\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-10-01 challenge.tex
took a pass on coqimpl
pWIP: redo §3 stlc figure

Inheritance and family polymorphism are mechanisms for
incrementally modifying self-referential definitions~\cite{cook1990inheritance}:
the implicit self parameterization is how they achieve late binding and
thus offer extensibility.
However, self reference could easily lead to divergence.
Divergence is not a concern for the design of ordinary OO or
functional languages, but it would mean logical inconsistency---and
hence unsoundness---for a language aimed at logical reasoning.
A family-polymorphism design must tame self reference to guarantee consistency.

@@

Interactive theorem proving and tactic programming are
part of the idiomatic programming experience in a proof assistant.
A language design integrating family polymorphism should be compatible
with these programming styles.
In particular, it should be possible in our tool to
incrementally navigate through vernacular commands and, moreover, construct proofs
using tactics with instant feedback on the proof state,
even in the middle of a family definition.
In addition, it should remain possible to perform computation on Gallina terms.

>>>>

Inheritance and family polymorphism are mechanisms for
incrementally modifying self-referential definitions~\cite{cook1990inheritance}:
the implicit self parameterization is how they achieve late binding and
thus offer extensibility.
However, self reference could easily lead to divergence.
Divergence is not a concern for the design of ordinary OO or
functional languages, but it would mean logical inconsistency---and
hence unsoundness---for a language aimed at logical reasoning!
A family-polymorphism design must tame self reference to guarantee consistency.

@@

Interactive theorem proving and tactic programming are
part of the idiomatic programming experience in a proof assistant.
A language design integrating family polymorphism should be compatible
with these programming styles.
In particular, it should be possible in our tool to
incrementally navigate through vernacular commands and, moreover, construct proofs
with common tactics while getting instant feedback on the proof state,
even in the middle of a family definition.
In addition, it should remain possible to perform computation in the extended Gallina language.

### 2022-10-01 lang-design.tex
pWIP: redo §3 stlc figure

We demonstrate key ingredients of our design as a plugin to the Coq proof assistant,
though we believe our design could be adapted to other proof assistants such as Lean.

@@

\textbf{Family and Defining Fields.} These are the core extensions we proposed to support \cref{fig:STLC-example}, and we adapt them into \cref{fig:STLC-example2}. We use "Family" and "EndFamily" to scope. "Field" is used to define new members in a family. "FieldClaim" is similar as "Field" but will invoke proof-mode in Coq and thus the user can use tactic programming to define new members.\footnote{We expect "FieldClaim" to be used in a \textit{proof-irrelevant} situation and thus whether overridable or not doesn't matter.} Both "Field" and "FieldClaim" are only allowed to refer to the former fields, to avoid possible inconsistency mentioned in \ref{chg:consistency}.
Nested families are also allowed.

@@

To resolve \ref{chg:consistency}, we make sure when inheriting or
overriding on "f" happens, the context of the field "f" is also
``inherited''. Thus the mentioned informal example in
\ref{chg:consistency} will not apply here, because the overriding "f" is
in a context where "self" is empty thus "self.g" doesn't
type-check. Thus, before we inherit "g", the dependee "f" needs to be
inherited.

@@

% % Talk about Final
% \textbf{Final as non-overridable.} Every field is by default
% overridable, reminding us of \ref{chg:definition-relevant-reasoning}.  To
% resolve it, we provide the keyword "Final" to allow the user to specify
% \textit{non-overridability} and give the user the opportunity to choose
% from overriding and exposing the definition.

% Take the field "c" of "Family I" in \cref{fig:plugin-example3} as an 
% example. We can see that the first attempt of proving "a = 1" fails.
% This should fail---because we want to retain the possibility
% to override "a" and inheriting "c" in the future, but they are
% conflicting, since "c" witnesses that "a" is fixed to "1". On the other
% hand, once we decorate \lstinline{a'} as final, the second attempt
% \lstinline{c : a' = 1 := eq_refl} will succeed. Contrary to "a",
% our \lstinline{a'} is not overridable in "Family I2", mediating the
% \ref{chg:definition-relevant-reasoning}.

@@

\textbf{Overridable Fields with dependency.}  Each field is by default non-overridable. By the discussion of \ref{chg:definition-relevant-reasoning}, we can simply reason and access the definition about the former fields in a later field, just like Field "a" and "b" in \cref{fig:plugin-example3} and we lose the ability to override.

@@

To resolve, we introduce overridable fields. Each overridable field will only expose its type instead of the concrete definition to the following default (non-overridable) fields. For example, the first failure in \cref{fig:plugin-example3} per the discussion of \ref{chg:definition-relevant-reasoning}: since overridability will break the inheritability of "c" and thus we don't allow such field to exist in the first place. Our plugin will only expose the fact that "a'" is an arbitrary natural number but hiding its concrete definition, and thus "eq_refl" will fail to type-check.

@@

To resolve \ref{chg:extensible-inductive-type}, our first step is to
\textit{decouple} the definition of recursor handlers and the checking
of exhaustiveness. The users will define recursive handlers for one
inductive type by aggregating the (function) handlers into one family.
Then
"FRecursor" will be used to construct a recursor by referring to the handler family.
\mintinline{Coq}{FRecursor recur about T motive f } "using"
"handlers" will generate a new field "recur" that is really doing computation
(using elimination principle of the inductive type "T") with motive "f". The handlers
are located in the family "handlers", similar to a large pattern-matching statement. This way, the reuse and the extension of the case handlers will be delegated by the inheritance of the family. Once recursor is created it will look like any other function fields(e.g. "subst" in \cref{fig:STLC-example2}).

@@

% We can also use "Family" and "SealFamily" to scope a (nested) family. The resulting family only support direct inheritance or "Override" (i.e. we can only override a sealed family as a whole together)---we will later explain this design decision and its usage, especially on resolving \ref{chg:definition-relevant-reasoning}. What's more, "Override" family can only apply to sealed family as well (not allowed on normal family).

% We can inherit a top-level family "parent" using \mintinline{Coq}{Family ... extends parent} as a new top-level family. As expected, "Override"
% will override a field from the parent. "Inherit" can be used to inherit
% nested families, inductive types, and fields. "Final" is used to claim that the decorated field is non-overridable, and later we will show how it is used to resolve \ref{chg:definition-relevant-reasoning}. The fields inside a family
% are order-sensitive because we have dependent types here. So "Inherit
% Until" is used if the user wants to inherit a range of fields.  "Extend Family" is used only inside a inheriting family. Thus nested
% family can only be inherited or extended together with its belonging family
% (i.e. ultimately top level family), thus our support of family polymorphism is
% incomplete.

We will explain the usage of "Closing Fact" and "MetaData" later---they are \textit{escape-hatches} to resolve \ref{chg:software-engineering} to some extent.
% This incomplete support can already be useful for extending functions.
% Take \cref{fig:plugin-example2} as example, if we want to extend boolean
% inductive type and its negation, we only need to specify the new
% constructor for boolean and how negation deals with the new constructor.
% As a result, the programmer does not have to duplicate code---otherwise
% the user will need to copy and paste the whole original negation
% definition and add a new pattern matching case.

>>>>

We demonstrate the key ingredients of our design as a plugin to the Coq proof assistant,
though we believe our design could be adapted to other proof assistants such as Lean.

@@

\textbf{Family and Defining Fields.} These are the core extensions we proposed to support \cref{fig:STLC-example}, and we adapt them into \cref{fig:STLC-example2}. We use "Family" and "EndFamily" to scope. "Field" is used to define new members in a family. "FieldClaim" is similar as "Field" but will invoke proof-mode in Coq and thus the user can use tactic programming to define new members.\footnote{We expect "FieldClaim" to be used in a \textit{proof-irrelevant} situation and thus whether overridable or not doesn't matter.} Both "Field" and "FieldClaim" are only allowed to refer to the former fields, to avoid possible inconsistency mentioned in \ref{chg:self-v-consistency}.
Nested families are also allowed.

@@

To resolve \ref{chg:self-v-consistency}, we make sure when inheriting or
overriding on "f" happens, the context of the field "f" is also
``inherited''. Thus the mentioned informal example in
\ref{chg:self-v-consistency} will not apply here, because the overriding "f" is
in a context where "self" is empty thus "self.g" doesn't
type-check. Thus, before we inherit "g", the dependee "f" needs to be
inherited.

@@

% % Talk about Final
% \textbf{Final as non-overridable.} Every field is by default
% overridable, reminding us of \ref{chg:override-v-dependent}.  To
% resolve it, we provide the keyword "Final" to allow the user to specify
% \textit{non-overridability} and give the user the opportunity to choose
% from overriding and exposing the definition.

% Take the field "c" of "Family I" in \cref{fig:plugin-example3} as an 
% example. We can see that the first attempt of proving "a = 1" fails.
% This should fail---because we want to retain the possibility
% to override "a" and inheriting "c" in the future, but they are
% conflicting, since "c" witnesses that "a" is fixed to "1". On the other
% hand, once we decorate \lstinline{a'} as final, the second attempt
% \lstinline{c : a' = 1 := eq_refl} will succeed. Contrary to "a",
% our \lstinline{a'} is not overridable in "Family I2", mediating the
% \ref{chg:override-v-dependent}.

@@

\textbf{Overridable Fields with dependency.}  Each field is by default non-overridable. By the discussion of \ref{chg:override-v-dependent}, we can simply reason and access the definition about the former fields in a later field, just like Field "a" and "b" in \cref{fig:plugin-example3} and we lose the ability to override.

@@

To resolve, we introduce overridable fields. Each overridable field will only expose its type instead of the concrete definition to the following default (non-overridable) fields. For example, the first failure in \cref{fig:plugin-example3} per the discussion of \ref{chg:override-v-dependent}: since overridability will break the inheritability of "c" and thus we don't allow such field to exist in the first place. Our plugin will only expose the fact that "a'" is an arbitrary natural number but hiding its concrete definition, and thus "eq_refl" will fail to type-check.

@@

To resolve \ref{chg:extensible-v-exhaustive}, our first step is to
\textit{decouple} the definition of recursor handlers and the checking
of exhaustiveness. The users will define recursive handlers for one
inductive type by aggregating the (function) handlers into one family.
Then
"FRecursor" will be used to construct a recursor by referring to the handler family.
\mintinline{Coq}{FRecursor recur about T motive f } "using"
"handlers" will generate a new field "recur" that is really doing computation
(using elimination principle of the inductive type "T") with motive "f". The handlers
are located in the family "handlers", similar to a large pattern-matching statement. This way, the reuse and the extension of the case handlers will be delegated by the inheritance of the family. Once recursor is created it will look like any other function fields(e.g. "subst" in \cref{fig:STLC-example2}).

@@

% We can also use "Family" and "SealFamily" to scope a (nested) family. The resulting family only support direct inheritance or "Override" (i.e. we can only override a sealed family as a whole together)---we will later explain this design decision and its usage, especially on resolving \ref{chg:override-v-dependent}. What's more, "Override" family can only apply to sealed family as well (not allowed on normal family).

% We can inherit a top-level family "parent" using \mintinline{Coq}{Family ... extends parent} as a new top-level family. As expected, "Override"
% will override a field from the parent. "Inherit" can be used to inherit
% nested families, inductive types, and fields. "Final" is used to claim that the decorated field is non-overridable, and later we will show how it is used to resolve \ref{chg:override-v-dependent}. The fields inside a family
% are order-sensitive because we have dependent types here. So "Inherit
% Until" is used if the user wants to inherit a range of fields.  "Extend Family" is used only inside a inheriting family. Thus nested
% family can only be inherited or extended together with its belonging family
% (i.e. ultimately top level family), thus our support of family polymorphism is
% incomplete.

We will explain the usage of "Closing Fact" and "MetaData" later---they are \textit{escape-hatches} to resolve \ref{chg:tooling} to some extent.
% This incomplete support can already be useful for extending functions.
% Take \cref{fig:plugin-example2} as example, if we want to extend boolean
% inductive type and its negation, we only need to specify the new
% constructor for boolean and how negation deals with the new constructor.
% As a result, the programmer does not have to duplicate code---otherwise
% the user will need to copy and paste the whole original negation
% definition and add a new pattern matching case.

### 2022-10-01 stlc-mechanized.tex
pWIP: redo §3 stlc figure

\begin{figure}
  \begin{minipage}[t]{0.47\linewidth}
\begin{lstlisting}[language=Coq,  escapeinside={@}{@}]
Family STLC.
 FInductive ty : Set :=
  | unit : ty
  | arrow : ty -> ty -> ty.
 FInductive term : Set := 
  | tm_var : id -> term 
  | tm_abs : id -> term -> term ...
 Family subst_internal.
  Final Field tm_var := ..
  Final Field tm_abs := ..
   ...
 EndFamily.
 FRecursor subst
  about tm motive ..
  using subst_internal.
 ...
 FInductive has_type 
  : context -> term -> type := ...
 FTheorem subst_lemma 
  about has_type 
  motive ...
 Proof.  ...
 Qed.
 FInductive step : term -> term -> Prop 
  := ...
(* ... and more, end with type safety *)
EndFamily.
\end{lstlisting}
  \end{minipage}
  \begin{minipage}[t]{0.47\linewidth}
\begin{lstlisting}[language=Coq,  escapeinside={@}{@}]
Family STLC_bool extends STLC.
 Extend FInductive ty : Set :=
  | bool : ty.

>>>>

\begin{figure}
\begin{minipage}{\textwidth}
\begin{multicols}{2}

\begin{lstlisting}
Family STLC.
FInductive ty : Set :=[@\hfill\codecomment{Types}@]
| ty_unit  : ty
| ty_arrow : ty -> ty -> ty.

FInductive tm : Set :=[@\hfill\codecomment{Terms}@]
| tm_var  : id -> tm
| tm_unit : tm
| tm_abs  : id -> tm -> tm
| tm_app  : tm -> tm -> tm.

Field env : Type := id -> option ty.
Field empty : env := \lambda _, None.

FInductive ofty : env -> tm -> ty -> Prop :=
| ot_var  : [@\dadada@][@\hfill\codecomment{Typing rules}@]
| ot_unit : [@\dadada@]
| ot_abs  : [@\dadada@]
| ot_app  : [@\dadada@].

FInductive value : tm -> Prop :=[@\hfill\codecomment{Value forms}@]
| v_unit : value tm_unit
| v_abs  : \forall x e, value (tm_abs x e).

FRecursion subst on tm[@\hfill\codecomment{Subst function}@] 
  motive \lambda(_ : tm), id -> tm -> tm.
Case tm_var := \lambda y x e,
  if (eqb x y) then e else (tm_var y)
Case tm_unit := \lambda x e, tm_unit
Case tm_abs := \lambda y e' IHe' x e, [@\dadada@]
Case tm_app := \lambda e1 IHe1 e2 IHe2 x e,
  tm_app (IHe1 x e) (IHe2 x e).
End subst.

FInductive step : tm -> tm -> Prop :=
| st_app1 : [@\dadada@][@\hfill\codecomment{Reduction rules}@]
| st_app2 : [@\dadada@]
| st_beta : \forall x e v, value v ->
  step (tm_app (tm_abs x e) v) (subst e x v).

FInduction subst_lemma on ofty[@\hfill\codecomment{Subst lemma}@]
  motive \lambda G' e T (_ : ofty G' e T),
  \forall G x e' T', G' = extend G x T' ->
  ofty empty e' T' -> ofty G (subst e x e') T.
Case ot_var.  [@\dadada@] Qed.
Case ot_unit. [@\dadada@] Qed.
Case ot_abs.  [@\dadada@] Qed.
Case ot_app.  [@\dadada@] Qed.
End subst_lemma.

End STLC.

Family STLC_fix extends STLC.

End STLC_fix.

\end{lstlisting}

\end{multicols}
\end{minipage}

\vspace{2em}

  \begin{minipage}[t]{0.47\linewidth}
\begin{lstlisting}[language=Coq,  escapeinside={@}{@}]
Family STLC.
 FInductive ty : Set :=
  | unit : ty
  | arrow : ty -> ty -> ty.
 FInductive term : Set := 
  | tm_var : id -> term 
  | tm_abs : id -> term -> term ...
 Family subst_internal.
  Final Field tm_var := ..
  Final Field tm_abs := ..
   ...
 EndFamily.
 FRecursor subst
  about tm motive ..
  using subst_internal.
 ...
 FInductive has_type 
  : context -> term -> type := ...
 FTheorem subst_lemma 
  about has_type 
  motive ...
 Proof.  ...
 Qed.
 FInductive step : term -> term -> Prop 
  := ...
(* ... and more, end with type safety *)
EndFamily.
\end{lstlisting}
  \end{minipage}
  \begin{minipage}[t]{0.47\linewidth}
\begin{lstlisting}[language=Coq,  escapeinside={@}{@}]
Family STLC_bool extends STLC.
 Extend FInductive ty : Set :=
  | bool : ty.
